---
title: "Cyclistic Bike Share"
author: "Chukwuemeka Collins Ochie"
date: "2023-03-27"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE, echo=FALSE, error=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, error = FALSE)
library(lubridate)
library(tidyverse)
library("janitor")

library(tidyr)

library("skimr")

library(ggplot2)
library(hms)

library(scales)
library("dplyr")

install.packages('forecast')
library(forecast)
```

## Data Wrangling Change Log

* Date Range: 2022-04-05 to 2023-03-07

April 05, 2022:
*Total rows: 5829084
*I raised the following concerns about the data with Lily Moreno ( The director of marketing):
(1) 850418 empty rows
(2) 511 rows with negative trip duration

* I was adviced by Lily Moreno that we could ignore rows with negative trip duration.
We could ignore empty rows with missing start_station_name and ignore station_id.
* And should focus on doing using the start_station_name to perform aggregate functions on date, start_station_name, member_casual and ridable_type.
* Cleaning reducted rows to 4978155

```{r}
##
## Import previous 12 months data
##
df1 <- read.csv("202203-divvy-tripdata.csv")
df2 <- read.csv('202204-divvy-tripdata.csv')
df3 <- read.csv('202205-divvy-tripdata.csv')
df4 <- read.csv('202206-divvy-tripdata.csv')
df5 <- read.csv('202207-divvy-tripdata.csv')
df6 <- read.csv('202208-divvy-tripdata.csv')
df7 <- read.csv('202209-divvy-publictripdata.csv')
df8 <- read.csv('202210-divvy-tripdata.csv')
df9 <- read.csv('202211-divvy-tripdata.csv')
df10 <- read.csv('202212-divvy-tripdata.csv')
df11 <- read.csv('202301-divvy-tripdata.csv')
df12 <- read.csv('202302-divvy-tripdata.csv')
```

 
```{r}
##
## Combine 12 data.frames into (1) data.frame
##
bike_rides <- rbind(df1, df2, df3, df4, df5, df6, df7, df8, df9, df10, df11, df12)
bike_rides <- janitor::remove_empty(bike_rides, which = c("cols"))
bike_rides <- janitor::remove_empty(bike_rides, which = c("rows"))
bike_rides <- bike_rides %>% filter(start_station_name !="")
```
```{r}
##
## Convert date/time stamp to date/time...
##
bike_rides$Ymd <- as.Date(bike_rides$started_at)
bike_rides$started_at <- lubridate::ymd_hms(bike_rides$started_at)
bike_rides$ended_at <- lubridate::ymd_hms(bike_rides$ended_at)

bike_rides$start_hour <- lubridate::hour(bike_rides$started_at)
bike_rides$end__hour <- lubridate::hour(bike_rides$ended_at)
```

```{r}
bike_rides$Hours <- difftime(bike_rides$ended_at, bike_rides$started_at, units = c("hours"))

bike_rides$Minutes <- difftime(bike_rides$ended_at, bike_rides$started_at, units = c("mins"))

bike_rides <- bike_rides %>% filter(Minutes >0)
```

```{r}
### Create summary data frame

bikesrides2 <- bike_rides %>% group_by(Weekly = floor_date(Ymd, "week"),start_hour) %>% summarise(Minutes = sum(Minutes), Mean = mean(Minutes), Median = median(Minutes), Max = max(Minutes), Min = min(Minutes), Count = n()) %>% ungroup()
```


```{r}
bikesrides2$CntMA <- forecast::ma(bikesrides2$Count, 28)
```

## Plot of rides by Date
### Summary stst: Counts

 *Summary of Hourly counts
```{r}
# Summary of Hourly Counts
summary(bikesrides2$Count)
```

 *Count of rides by hour
```{r}
xtabs(bikesrides2$Count~bikesrides2$start_hour)
```

```{r}
bikesrides2$Monthly <- lubridate::month(bikesrides2$Weekly)
```


```{r}
bikesrides2 %>% ggplot() + geom_col(aes(x=Weekly, y=Count)) + 
  scale_y_continuous(labels = comma) + 
  labs(title = "Count of Rides per Day",
       subtitle = "(Bases on 28 day moving average", 
       y="average rides per day")
```

```{r}
bikesrides2 %>% ggplot() + geom_col(aes(x=start_hour, y=Count)) + 
  scale_y_continuous(labels = comma) + 
  labs(title = "Count of rides by Hours", y="Rides per Hour")
```
## Count of Rides by Bike Type 
#### Summary of Bike Types
```{r}
bikestype <- bike_rides %>% 
  group_by(member_casual, rideable_type, Weekly = 
  floor_date(Ymd, "week")) %>% 
                   summarise(Minutes = sum(Minutes),
                     Mean = mean(Minutes),
                     Median = median(Minutes),
                     Max = max(Minutes),
                     Min = min(Minutes), 
                     Count = n()
                   ) %>% ungroup()
```

*Count by Bike Type(Total by week)



```{r}
table(bikestype$rideable_type)
```


```{r}
ggplot(bikestype) + 
  geom_area(aes(x=Weekly, y=Count, fill=rideable_type))
```


```{r}
ggplot(bikestype) + 
  geom_area(aes(x=Weekly, y=Count, fill=rideable_type)) + 
  scale_y_continuous(labels = comma) + 
  labs(title="Count of Rides by Bike Type",subtitle = "(for a 12 month period ending 2023-mar -06)")

```

```{r}
ggplot(bikestype) + 
  geom_area(aes(x=Weekly, y=Count, fill=rideable_type)) + 
  facet_wrap(~rideable_type, scales = "free_y")
```

```{r}
ggplot(bikestype) + 
         geom_col(aes(x=Weekly, y=Count, fill=member_casual)) + 
         scale_y_continuous(labels = comma) + 
         labs(title="Count of Rides by Rider Type")
```

```{r}
ggplot(bikestype) + 
  geom_col(aes(x=Weekly, y=Count, col=member_casual)) + 
  scale_y_continuous(labels = comma) + 
  labs(title="Count of Rides by rider Type")
```

```{r}
bike_rides %>% count(start_station_name, sort = TRUE) %>% 
  top_n(20) %>% ggplot() + geom_col(aes(x=reorder(start_station_name, n), y=n)) + 
  coord_flip() + labs(title = "Top 20 start stations by Ride Count", 
                      y = "station Name",x="Count of Rides") + 
  scale_y_continuous(labels = comma)
```

```{r}
ggplot(bikestype) + 
  geom_area(aes(x=Weekly, y=Count, fill=rideable_type)) + 
  scale_y_continuous(labels = comma) + 
  labs(title="Count of Rides by Bike Type")
```

```{r}
ggplot(bikestype) + 
  geom_col(aes(x=Weekly, y=Minutes)) + 
  scale_y_continuous(labels = comma) +
  facet_wrap(~rideable_type) + 
  labs(title = "Total Ride Minutes by Week")
```

```{r}
ggplot(bikestype, aes(x=Weekly, y=Minutes, fill=rideable_type)) + 
  geom_col(stat = "identity", position = position_dodge(),
            alpha = 0.75) + 
  scale_y_continuous(labels = comma) + 
  labs(title = "Rides Minutes by Bike Type and Week", 
       y="Bike trip in Minutes")

```

